---
title: "One Run of the Agent Search Model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{One Run of the Agent Search Model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
library(scerrModel)
lgr::get_logger("scerrModel")$set_threshold("fatal")
```

## Scope

The package separates a simulation into two steps. First,
`simulate_reality()` generates the true effect and the faults in a study.
Second, the internal function `run_agent_model()` represents one agent deciding
whether to inspect the study one unit at a time. `simulate_literature()` calls
both functions for every requested study.

This vignette describes the second step: one call to `run_agent_model()`. The
objective reality is held fixed, and the focus is the agent's changing belief
about the total number of faults. The notation below follows the implementation
closely enough that each equation can be matched to a stored value.

## Inputs to one run

A run receives study, agent, and objective-reality inputs.

| Group | Code names | Meaning |
|---|---|---|
| Study | `N`, `resources`, `cost`, `benefit` | Number of units, initial search budget, cost of one inspection, and benefit of finding a fault |
| Prior on total faults | `sbj_prob_alpha`, `sbj_prob_beta` | Shape parameters of the beta-binomial prior for the total number of faults |
| Prior on the true effect | `sbj_effect_mu`, `sbj_effect_sigma2` | Mean and variance of the agent's normal prior for the true effect |
| Prior on error sizes | `sbj_error_mu`, `sbj_error_kappa`, `sbj_error_var_alpha`, `sbj_error_var_beta` | Normal--Inverse-Gamma parameters for error sizes |
| Objective reality | `true_effect`, `faults`, `error_sizes` | The simulated truth that the agent encounters during the search |

Let the study contain $N$ units. Unit $j$ has a logical fault status $z_j$.
It is `TRUE` when the unit has a fault and `FALSE` otherwise. Treating these
values as one and zero, respectively,

$$
K = \sum_{j=1}^{N} z_j
$$

is the true total number of faults. In the implementation, `error_sizes` has
length $N$: its $j$-th entry is zero for a non-faulty unit and is the error
contribution $e_j$ for a faulty unit. The initially observed effect is

$$
y_0 = \theta + \sum_{j=1}^{N} e_j,
$$

where $\theta$ is `true_effect`. `simulate_literature()` stores `true_K` in its
output but does not pass it to `run_agent_model()`. The agent learns from the
inspected fault values, observed error sizes, and current observed effect.

## The prior over the number of faults

The agent starts with a beta-binomial prior:

$$
K \sim \operatorname{BetaBinomial}(N,\alpha_K,\beta_K),
$$

where `sbj_prob_alpha` is $\alpha_K$ and `sbj_prob_beta` is
$\beta_K$. For $k=0,\ldots,N$,

$$
P(K=k)
=
{N\choose k}
\frac{B(k+\alpha_K,N-k+\beta_K)}{B(\alpha_K,\beta_K)}.
$$

This distribution is stored as the first row of `history$fault_posteriors`. Its
columns represent $K=0,\ldots,N$.

## State and round indexing

Before inspection round $i\in\{1,\ldots,N\}$, the agent has completed
$i-1$ inspections. Define:

- $n_{\mathrm{found},i-1}$: number of faults found in those inspections;
- $x_1,\ldots,x_{n_{\mathrm{found},i-1}}$: their observed error sizes;
- $y_{i-1}$: current observed effect after correcting those errors;
- $R_{i-1}$: remaining resources;
- $\pi_{i-1}(k)$: current posterior probability that $K=k$.

The code uses the following indexing convention:

- row 1 of `fault_posteriors` is $\pi_0$, the prior;
- row $i+1$ is $\pi_i$, after inspection $i$;
- `next_fault_belief[round]` and `eu_criterion[round]` describe the decision made before
  inspection $i$;
- `rounds_completed` is the number of inspections actually completed.

## Decision before an inspection

The current posterior assigns probabilities to candidate totals $k$. Given
that $n_{\mathrm{found},i-1}$ faults have already been found, the number still
undiscovered under candidate $k$ is $k-n_{\mathrm{found},i-1}$. The
implementation first normalizes the posterior over feasible candidates
$k\geq n_{\mathrm{found},i-1}$, then calculates

$$
\operatorname{E}[K-n_{\mathrm{found},i-1}\mid\mathcal D_{i-1}]
=
\sum_{k=n_{\mathrm{found},i-1}}^N
(k-n_{\mathrm{found},i-1})\,\pi_{i-1}(k).
$$

There are $N-i+1$ units left to inspect. The subjective probability that the
next unit contains a fault is therefore

$$
p_i
=
\frac{\operatorname{E}[K-n_{\mathrm{found},i-1}\mid\mathcal D_{i-1}]}
     {N-i+1}.
$$

This is `next_fault_belief[round]`. The expected-utility criterion is

$$
c_i = p_i\frac{\text{benefit}}{\text{cost}},
$$

stored as `eu_criterion[round]`. The agent continues only if $c_i\geq1$ and
the remaining resources are at least `cost`. If it continues, the cost is
deducted before the unit is inspected. Otherwise, the run stops without
completing round $i$.

The code checks expected utility before resources. If both conditions fail at
the same decision, the recorded reason is therefore `"Expected utility too
low"`.

## Observation and correction

If unit $i$ is faulty, the fault count increases and its error is removed from
the observed effect:

$$
n_{\mathrm{found},i}=n_{\mathrm{found},i-1}+1,
\qquad
y_i=y_{i-1}-e_i.
$$

For a non-faulty unit, $n_{\mathrm{found},i}=n_{\mathrm{found},i-1}$ and
$y_i=y_{i-1}$. The order of units is fixed by `faults`; the agent does not choose which unit to inspect
next.

## Posterior update after an inspection

After round $i$, the implementation recomputes the posterior from the
original beta-binomial prior and the cumulative observations. For each feasible
candidate $k\in\{n_{\mathrm{found},i},\ldots,N\}$, the unnormalized weight is

$$
w_i(k)
=
P(K=k)\,
L_{\mathrm{count},i}(k)\,
L_{\mathrm{size},i}\,
L_{\mathrm{effect},i}(k).
$$

The normalized posterior is

$$
\pi_i(k)=\frac{w_i(k)}{\sum_{m=0}^N w_i(m)}.
$$

The model includes three likelihood terms, although only two depend on the
candidate count.

### Fault-count likelihood

After finding $n_{\mathrm{found},i}$ faults in $i$ inspected units, the count likelihood for
candidate $K=k$ is hypergeometric:

$$
L_{\mathrm{count},i}(k)
=
P(n_{\mathrm{found},i}\mid K=k,i,N)
=
\frac{{k\choose n_{\mathrm{found},i}}{N-k\choose i-n_{\mathrm{found},i}}}{{N\choose i}}.
$$

This rules out totals that are incompatible with the inspected units.

### Marginal likelihood of observed error sizes

The implementation evaluates the closed-form marginal likelihood of the
observed error sizes under the Normal--Inverse-Gamma prior. This is
$L_{\mathrm{size},i}$. It uses `sbj_error_kappa`, but it is identical for every
candidate total $k$ and therefore cancels during normalization. Consequently,
`sbj_error_kappa` is currently a dead parameter. The calculation is retained
pending a correction to the error-belief model.

### Likelihood of the current observed effect

For candidate $K=k$, let

$$
r_i(k)=k-n_{\mathrm{found},i}
$$

be the candidate number of undiscovered faults. The implementation replaces
the error variance by the mean of its inverse-gamma prior,

$$
\sigma_e^{2*}=\frac{\beta_e}{\alpha_e-1},
$$

which requires $\alpha_e>1$. It then uses

$$
y_i\mid K=k
\sim
\mathcal N\!\left(
\mu_\theta+r_i(k)m_e,
\sigma_\theta^2+r_i(k)\sigma_e^{2*}
\right).
$$

Thus,

$$
L_{\mathrm{effect},i}(k)
=
\phi\!\left(
y_i;
\mu_\theta+r_i(k)m_e,
\sigma_\theta^2+r_i(k)\sigma_e^{2*}
\right).
$$

When $r_i(k)=0$, this correctly reduces to the agent's prior density for the
true effect. This part of the implementation is a plug-in normal approximation,
not the full posterior-predictive distribution implied by the
Normal--Inverse-Gamma prior.

### Numerical implementation

The implementation adds the log prior and all three log likelihoods, subtracts
their maximum, and exponentiates before normalization. This log-sum-exp
calculation retains posterior information when the likelihoods are too small
to represent on the probability scale. Learning the error-size distribution
would require the observed errors to affect a posterior-predictive likelihood
for the remaining errors.

## A deterministic worked run

The following objective reality has five units and two faults. The first fault
inflates the observed effect by 0.4; the second changes it by -0.2. The true
effect is 0.5, so the initially observed effect is $0.5+0.4-0.2=0.7$.

`run_agent_model()` is internal rather than part of the public user API. It is
called directly here only to make every input to the worked run visible.

```{r worked-run}
search_args <- list(
  study_id = "worked-study",
  agent_id = "worked-agent",
  N = 5,
  benefit = 10,
  cost = 1,
  resources = 5,
  sbj_prob_alpha = 2,
  sbj_prob_beta = 3,
  sbj_effect_mu = 0.5,
  sbj_effect_sigma2 = 0.25,
  sbj_error_mu = 0,
  sbj_error_kappa = 1,
  sbj_error_var_alpha = 3,
  sbj_error_var_beta = 1,
  true_effect = 0.5,
  faults = c(FALSE, TRUE, FALSE, TRUE, FALSE),
  error_sizes = c(0, 0.4, 0, -0.2, 0)
)

run <- do.call(scerrModel:::run_agent_model, search_args)
```

The table below shows the state after each completed round. The decision
columns on row $i$ refer to the proposed next round $i+1$. There is no next
decision after round 5.

```{r worked-state-table}
N <- search_args$N
faults_support <- 0:N
posterior <- run$history$fault_posteriors

state <- data.frame(
  round = 0:N,
  observation = c("start", search_args$faults),
  corrected_error = c(0, search_args$error_sizes),
  faults_found = c(0, cumsum(search_args$faults)),
  observed_effect = search_args$true_effect + sum(search_args$error_sizes) -
    c(0, cumsum(search_args$error_sizes)),
  resources_left = search_args$resources - (0:N) * search_args$cost,
  posterior_mean_K = as.vector(posterior %*% faults_support),
  next_fault_belief = c(unname(run$history$next_fault_belief), NA_real_),
  next_eu_criterion = c(unname(run$history$eu_criterion), NA_real_)
)

knitr::kable(state, digits = 3)
```

The prior mean is 2. After the first non-fault, lower values of $K$ receive
more posterior mass. Finding a fault in round 2 removes $K=0$ from the support.
The same pattern repeats in rounds 3 and 4. After all five units have been
inspected, the hypergeometric likelihood identifies $K=2$ with probability 1.

The complete posterior matrix makes this progression explicit:

```{r worked-posterior-table}
posterior_table <- data.frame(
  round = 0:N,
  round(posterior, 3),
  check.names = FALSE
)
names(posterior_table) <- c("Round", paste0("K = ", faults_support))
knitr::kable(posterior_table)
```

All expected-utility criteria in this run exceed 1, and resources cover all
five inspections. The search therefore completes:

```{r worked-stop-table}
stop_table <- data.frame(
  quantity = names(run$stop_conditions),
  value = unname(unlist(run$stop_conditions)),
  row.names = NULL
)
knitr::kable(stop_table)
```

## Output interpretation

`run_agent_model()` returns two nested lists:

- `stop_conditions` records how many inspections were completed, why the run
  stopped, remaining resources, the final observed effect, and the reported
  number of discovered faults;
- `history` contains the posterior matrix and the decision quantities for each
  possible round.

For a run that stops early, posterior rows after the last completed inspection
remain `NA`. The preallocated `next_fault_belief` and `eu_criterion` entries after
the stopping decision remain zero; those zeros are placeholders, not decisions
the agent actually made.

The reported number of discovered faults is the maintained count $n_{\mathrm{found}}$,
independently of the size of each error.
