---
title: "One Run of the Agent Search Model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{One Run of the Agent Search Model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
library(scerrModel)
lgr::get_logger("scerrModel")$set_threshold("fatal")
```

## Scope

The package separates a simulation into two steps. First,
`simulate_obj_reality()` generates the true effect and the faults in a study.
Second, the internal function `run_agent_model()` represents one agent deciding
whether to inspect the study one unit at a time. `simulate_literature()` calls
both functions for every requested study.

This vignette describes the second step: one call to `run_agent_model()`. The
objective reality is held fixed, and the focus is the agent's changing belief
about the total number of faults. The notation below follows the implementation
closely enough that each equation can be matched to a stored value.

## Inputs to one run

A run receives study, agent, and objective-reality inputs.

| Group | Code names | Meaning |
|---|---|---|
| Study | `N`, `resources`, `cost`, `benefit` | Number of units, initial search budget, cost of one inspection, and benefit of finding a fault |
| Prior on total faults | `sbj_prob_alpha`, `sbj_prob_beta` | Shape parameters of the beta-binomial prior for the total number of faults |
| Prior on the true effect | `sbj_effect_mu`, `sbj_effect_sigma2` | Mean and variance of the agent's normal prior for the true effect |
| Prior on error sizes | `sbj_error_mu`, `sbj_error_kappa`, `sbj_error_var_alpha`, `sbj_error_var_beta` | Normal--Inverse-Gamma parameters for error sizes |
| Objective reality | `true_effect`, `true_K`, `fault_indicators`, `error_sizes` | The simulated truth that the agent encounters during the search |

Let the study contain $N$ units. Unit $j$ has a fault indicator
$z_j\in\{0,1\}$, and

$$
K = \sum_{j=1}^{N} z_j
$$

is the true total number of faults. In the implementation, `error_sizes` has
length $N$: its $j$-th entry is zero for a non-faulty unit and is the error
contribution $e_j$ for a faulty unit. The initially observed effect is

$$
y_0 = \theta + \sum_{j=1}^{N} e_j,
$$

where $\theta$ is `true_effect`. The agent does not use `true_K` when updating
its beliefs. It learns from the inspected fault indicators, observed error
sizes, and current observed effect.

## The prior over the number of faults

The agent starts with a beta-binomial prior:

$$
K \sim \operatorname{BetaBinomial}(N,\alpha_K,\beta_K),
$$

where `sbj_prob_alpha` is $\alpha_K$ and `sbj_prob_beta` is
$\beta_K$. For $k=0,\ldots,N$,

$$
P(K=k)
=
{N\choose k}
\frac{B(k+\alpha_K,N-k+\beta_K)}{B(\alpha_K,\beta_K)}.
$$

This distribution is stored as the first row of `history$posterior_K`. Its
columns represent $K=0,\ldots,N$.

## State and round indexing

Before inspection round $i\in\{1,\ldots,N\}$, the agent has completed
$i-1$ inspections. Define:

- $b_{i-1}$: number of faults found in those inspections;
- $x_1,\ldots,x_{b_{i-1}}$: their observed error sizes;
- $y_{i-1}$: current observed effect after correcting those errors;
- $R_{i-1}$: remaining resources;
- $\pi_{i-1}(k)$: current posterior probability that $K=k$.

The code uses the following indexing convention:

- row 1 of `posterior_K` is $\pi_0$, the prior;
- row $i+1$ is $\pi_i$, after inspection $i$;
- `fault_belief[i]` and `eu_criterion[i]` describe the decision made before
  inspection $i$;
- `stopped_in_round` is the number of inspections actually completed.

## Decision before an inspection

The current posterior assigns probabilities to candidate totals $k$. Given
that $b_{i-1}$ faults have already been found, the number still undiscovered
under candidate $k$ is $k-b_{i-1}$. The implementation first normalizes the
posterior over feasible candidates $k\geq b_{i-1}$, then calculates

$$
\operatorname{E}[K-b_{i-1}\mid\mathcal D_{i-1}]
=
\sum_{k=b_{i-1}}^N
(k-b_{i-1})\,\pi_{i-1}(k).
$$

There are $N-i+1$ units left to inspect. The subjective probability that the
next unit contains a fault is therefore

$$
p_i
=
\frac{\operatorname{E}[K-b_{i-1}\mid\mathcal D_{i-1}]}
     {N-i+1}.
$$

This is `fault_belief[i]`. The expected-utility criterion is

$$
c_i = p_i\frac{\text{benefit}}{\text{cost}},
$$

stored as `eu_criterion[i]`. The agent continues only if $c_i\geq1$ and
the remaining resources are at least `cost`. If it continues, the cost is
deducted before the unit is inspected. Otherwise, the run stops without
completing round $i$.

The code checks expected utility before resources. If both conditions fail at
the same decision, the recorded reason is therefore `"Expected utility too
low"`.

## Observation and correction

If unit $i$ is faulty, the fault count increases and its error is removed from
the observed effect:

$$
b_i=b_{i-1}+1,
\qquad
y_i=y_{i-1}-e_i.
$$

For a non-faulty unit, $b_i=b_{i-1}$ and $y_i=y_{i-1}$. The order of units
is fixed by `fault_indicators`; the agent does not choose which unit to inspect
next.

## Posterior update after an inspection

After round $i$, the implementation recomputes the posterior from the
original beta-binomial prior and the cumulative observations. For each feasible
candidate $k\in\{b_i,\ldots,N\}$, the unnormalized weight is

$$
w_i(k)
=
P(K=k)\,
L_{\mathrm{count},i}(k)\,
L_{\mathrm{size},i}\,
L_{\mathrm{effect},i}(k).
$$

The normalized posterior is

$$
\pi_i(k)=\frac{w_i(k)}{\sum_{m=0}^N w_i(m)}.
$$

The three likelihood components are described next.

### Fault-count likelihood

After finding $b_i$ faults in $i$ inspected units, the count likelihood for
candidate $K=k$ is hypergeometric:

$$
L_{\mathrm{count},i}(k)
=
P(b_i\mid K=k,i,N)
=
\frac{{k\choose b_i}{N-k\choose i-b_i}}{{N\choose i}}.
$$

This rules out totals that are incompatible with the inspected units.

### Marginal likelihood of observed error sizes

The agent's error-size prior is parameterized as

$$
\mu_e^*\mid\sigma_e^2
\sim
\mathcal N\!\left(m_e,\frac{\sigma_e^2}{\kappa_0}\right),
\qquad
\sigma_e^2\sim\operatorname{InvGamma}(\alpha_e,\beta_e),
$$

where $m_e$, $\kappa_0$, $\alpha_e$, and $\beta_e$ correspond to the
four `sbj_error_*` parameters. For $b=b_i>0$ observed error sizes with mean
$\bar x$, the conjugate update used to calculate their marginal likelihood is

$$
\begin{aligned}
\kappa_b &= \kappa_0+b,\\
\alpha_b &= \alpha_e+\frac b2,\\
\beta_b &= \beta_e
+\frac12\sum_{j=1}^{b}(x_j-\bar x)^2
+\frac{\kappa_0b(\bar x-m_e)^2}{2\kappa_b}.
\end{aligned}
$$

The implemented marginal likelihood is

$$
L_{\mathrm{size},i}
=
\frac{\Gamma(\alpha_b)}{\Gamma(\alpha_e)}
\frac{\beta_e^{\alpha_e}}{\beta_b^{\alpha_b}}
\sqrt{\frac{\kappa_0}{\kappa_b}}
(2\pi)^{-b/2}.
$$

When no error sizes have been observed, this component is 1.

Importantly, $L_{\mathrm{size},i}$ does not depend on candidate $k$. In
exact arithmetic it therefore cancels from the normalized posterior over $K$.
As currently implemented, the updated Normal--Inverse-Gamma parameters are not
used in the remaining-effect likelihood below. Consequently,
`sbj_error_kappa` does not change the normalized posterior except through
finite-precision effects. A model intended to learn the error-size distribution
from observed errors would instead need a posterior-predictive likelihood for
the remaining errors.

### Likelihood of the current observed effect

For candidate $K=k$, let

$$
r_i(k)=k-b_i
$$

be the candidate number of undiscovered faults. The implementation replaces
the error variance by the mean of its inverse-gamma prior,

$$
\sigma_e^{2*}=\frac{\beta_e}{\alpha_e-1},
$$

which requires $\alpha_e>1$. It then uses

$$
y_i\mid K=k
\sim
\mathcal N\!\left(
\mu_\theta+r_i(k)m_e,
\sigma_\theta^2+r_i(k)\sigma_e^{2*}
\right).
$$

Thus,

$$
L_{\mathrm{effect},i}(k)
=
\phi\!\left(
y_i;
\mu_\theta+r_i(k)m_e,
\sigma_\theta^2+r_i(k)\sigma_e^{2*}
\right).
$$

When $r_i(k)=0$, this correctly reduces to the agent's prior density for the
true effect. This part of the implementation is a plug-in normal approximation,
not the full posterior-predictive distribution implied by the
Normal--Inverse-Gamma prior.

### Numerical implementation

The current code forms `w_i(k)` on the probability scale and then divides by
its sum. If every computed weight is zero, it stores an all-zero posterior row.
For large studies or very small likelihoods, a log-scale implementation with a
log-sum-exp normalization would be more stable. The candidate-independent
`L_size` factor can also be omitted when the target is only the posterior over
$K$.

## A deterministic worked run

The following objective reality has five units and two faults. The first fault
inflates the observed effect by 0.4; the second changes it by -0.2. The true
effect is 0.5, so the initially observed effect is $0.5+0.4-0.2=0.7$.

`run_agent_model()` is internal rather than part of the public user API. It is
called directly here only to make every input to the worked run visible.

```{r worked-run}
model_args <- list(
  study_id = "worked-study",
  agent_id = "worked-agent",
  N = 5,
  benefit = 10,
  cost = 1,
  resources = 5,
  sbj_prob_alpha = 2,
  sbj_prob_beta = 3,
  sbj_effect_mu = 0.5,
  sbj_effect_sigma2 = 0.25,
  sbj_error_mu = 0,
  sbj_error_kappa = 1,
  sbj_error_var_alpha = 3,
  sbj_error_var_beta = 1,
  true_effect = 0.5,
  true_K = 2,
  fault_indicators = c("no_fault", "fault", "no_fault", "fault", "no_fault"),
  error_sizes = c(0, 0.4, 0, -0.2, 0)
)

run <- do.call(scerrModel:::run_agent_model, model_args)
```

The table below shows the state after each completed round. The decision
columns on row $i$ refer to the proposed next round $i+1$. There is no next
decision after round 5.

```{r worked-state-table}
N <- model_args$N
K_values <- 0:N
posterior <- run$history$posterior_K

state <- data.frame(
  round = 0:N,
  observation = c("start", model_args$fault_indicators),
  corrected_error = c(0, model_args$error_sizes),
  faults_found = c(0, cumsum(model_args$fault_indicators == "fault")),
  observed_effect = model_args$true_effect + sum(model_args$error_sizes) -
    c(0, cumsum(model_args$error_sizes)),
  resources = model_args$resources - (0:N) * model_args$cost,
  posterior_mean_K = as.vector(posterior %*% K_values),
  next_fault_probability = c(unname(run$history$fault_belief), NA_real_),
  next_eu_criterion = c(unname(run$history$eu_criterion), NA_real_)
)

knitr::kable(state, digits = 3)
```

The prior mean is 2. After the first non-fault, lower values of $K$ receive
more posterior mass. Finding a fault in round 2 removes $K=0$ from the support.
The same pattern repeats in rounds 3 and 4. After all five units have been
inspected, the hypergeometric likelihood identifies $K=2$ with probability 1.

The complete posterior matrix makes this progression explicit:

```{r worked-posterior-table}
posterior_table <- data.frame(
  round = 0:N,
  round(posterior, 3),
  check.names = FALSE
)
names(posterior_table) <- c("Round", paste0("K = ", K_values))
knitr::kable(posterior_table)
```

All expected-utility criteria in this run exceed 1, and resources cover all
five inspections. The search therefore completes:

```{r worked-stop-table}
stop_table <- data.frame(
  quantity = names(run$stop_conditions),
  value = unname(unlist(run$stop_conditions)),
  row.names = NULL
)
knitr::kable(stop_table)
```

## Output interpretation

`run_agent_model()` returns two nested lists:

- `stop_conditions` records how many inspections were completed, why the run
  stopped, remaining resources, the final observed effect, and the reported
  number of discovered faults;
- `history` contains the posterior matrix and the decision quantities for each
  possible round.

For a run that stops early, posterior rows after the last completed inspection
remain `NA`. The preallocated `fault_belief` and `eu_criterion` entries after
the stopping decision remain zero; those zeros are placeholders, not decisions
the agent actually made.

Conceptually, the number of discovered faults is $b$. The current code reports
`sum(obs_error_sizes != 0)`, which differs from $b$ if a discovered fault has
an error contribution exactly equal to zero. That distinction is relevant when
using this vignette as a specification against which to compare the
implementation.
